SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-29 abs.tex
Gino's first pass, modifying the abs, intro, design_arch, and design_impl

\begin{abstract}
Hardware prefetching has been introduced in modern processors as a way to hide
cache latencies.  An efficient prefetcher should be able to identify complex
memory access patterns during program execution. This ability enables the
prefetcher to read a block ahead of its demand access, potentially saving a
cache miss. Accurately identifying the right blocks to prefetch is essential
to achieving high performance from the prefetcher.

In this paper, we introduce Perceptron-based Prefetch Filtering to help make
this prefetching decision accurately. The perceptron layer acts as a check 
to filter out the unnecessary prefetches recommended by the underlying
prefetcher.  We have also explored a range of features that can be used to
train the perceptron layer.  Our results show that perceptron-based filtering
improves performance on the memory intensive subset of the SPEC 2017 benchmark
suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
to a state-of-the art prefetcher.  We also demonstrate that the performance
gained from using our efficient filter scales better with increasing numbers
of cores.

>>>>

\begin{abstract}
Hardware prefetching has been introduced in modern processors as a way to hide
cache latencies.  An efficient prefetcher should identify complex
memory access patterns during program execution. This ability enables the
prefetcher to read a block ahead of its demand access, potentially preventing a
cache miss. Accurately identifying the right blocks to prefetch is essential
to achieving high performance from the prefetcher.

In this paper, we introduce Perceptron-based Prefetch Filtering to help make
 prefetching decisions accurately. The perceptron layer acts as a check 
to filter out the unnecessary prefetches recommended by the underlying
prefetcher.  We have also explored a range of features that can be used to
train the perceptron layer.  Our results show that perceptron-based filtering
improves performance on the memory intensive subset of the SPEC 2017 benchmark
suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
to a state-of-the art prefetcher.  We also demonstrate that the performance
gained from using our efficient filter continues to scale as the number of cores sharing a 
last level cache increases.

### 2018-11-29 design_arch.tex
Gino's first pass, modifying the abs, intro, design_arch, and design_impl

In this section, we talk about the the general structure of the 
Perceptron-Based Prefetch Filtering (PPF) for an underlying prefetcher.  
PPF is a generalised prefetch filtering mechanism and can be adapted to 
most prefetchers with minimal modifications.

\subsection{The Perceptron Filter}
\label{Arch-Perceptron}
The Perceptron Filter is organised as tables of perceptron weights, 
with a separate table for each feature.  
The feature value is used to index into the corresponding table.
This indexing is done using different numbers of bits for each feature.  
At most 12 bits of a feature are used to index into the
weights table.  Certain features require more resolution power
\textit{i.e.}, the full 12 bits of indexing.  On the other hand, some
features requiring lower resolution require as few as 7 bits of
indexing.  The variable indexing was determined by studying the
features and fine-tuned empirically so as to achieve a good accuracy
vs area trade-off.  Exact details are mentioned under ``Area Overheads''
in Section~\ref{Method-Overheads}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15.  
Our proposed design uses nine features, hence there are nine perceptron 
weight tables. As program execution begins, all the weights are initialized to 0.

\textbf{Inference}\newline In the ChampSim simulator, the prefetching method is 
invoked on every L2 Cache demand access.  At that point, the base prefetcher
has the option to do or not do a prefetch -- if it does, then it has a choice
in how many cache lines to prefetch.  The suggested prefetches can be either
in the L2 Cache or the last-level cache.  The prefetcher decides this
placement, and it is usually done based on an internal confidence mechanism.

When the base prefetcher is triggered, it starts suggesting candidates for
prefetching.  All these recommended prefetches have to be tested through the perceptron
filter to qualify for final prefetching.

The perceptron looks at the microarchitectural state, {\em i.e.} the features,
at that instant.  Each feature is hashed to form an index into a table of up to 
4096 entries dedicated for that feature.  
Feature \#1 indexes into a particular entry in table \#1 and so on.

Once all the weights are retrieved, they are summed.  The sum is thresholded
based on the preset threshold PERC\_ THRESHOLD\_LO.  Only the prefetch
candidates with perceptron sum higher that the threshold qualify for
prefetching.  The prefetches that qualify through the perceptron 
stage are recorded in the Prefetch Table.  In PPF, all the metadata that is required
to recreate the features is also stored in the Prefetch Table.  At a later
stage, when the feedback of the current prefetched line is available, the
stored data can be used to train the perceptron.

The perceptron sum is also used to decide L2 Cache vs Last Level Cache
placement of the prefetch cache line.  All prefetch candidates qualified till
this stage are thresholded against PERC\_THRESHOLD\_HI to decide the fill
level. The two thresholds: PERC\_THRESHOLD\_LO and PERC \_THRESHOLD\_HI are
empirically set.

The perceptron sum can also be seen as the sum of the individual contribution
per feature.  The value of each contribution corresponds to the extent of
confidence of the final output with that given feature.  By summing the
individual contributions, the final perceptron sum denotes the overall
confidence for that prefetch suggestion.  By thresholding the perceptron sum
on two different thresholds, we are dividing the confidence scale of the
prefetch suggestion into three bins.  The first bin corresponds to the lowest
confidence leads to prefetch candidate being rejected.  The next bin
corresponds to prefetches with a moderate confidence level.  Such prefetches
are directed to fill the bigger Last Level Cache and potentially not pollute
the more scarce L2 Cache.  The highest confidence prefetches are filled in the
L2 Cache.

In addition to the Prefetch Table mentioned above, PPF also maintains a
``Reject Table.''  The Reject Table is a 1024-entry deep direct-mapped
table.  If a prefetch suggestion is rejected by the perceptron layer, it is
logged into the Reject Table.  The filer is used to train the perceptron to
avoid false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

@@

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 Cache.  This action triggers training of the base prefetcher as well as the
perceptron layer.  Training the perceptrons involves restoring
the metadata that was stored in Prefetch Table or the Reject Table.  This is
done by indexing into the corresponding tables using the cache line address
of the block being used to train. (10-bits indexing + 6-bits tag matching).
Once the state of the program at the time of prefetching is available, it is
used to index into the perceptron weights table.

If a demand access block that triggers the training was tagged as a valid
prefetch in the Prefetch Table, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 
if the predicted sum does not cross a pre-defined threshold. These training 
thresholds are introduced to limit the degree of update on correct predictions.
Doing this helps avoid overfitting of the perceptron weights to the given program 
behaviour. These thresholds are referred to as POS\_UPDATE\_THRESHOLD ($\theta_p$) and 
NEG\_UPDATE\_THRESHOLD ($\theta_n$),  respectively for the positive and 
negative values of training saturation.

If a cache block
eviction led to training and the corresponding valid entry was found in the
Prefetch Table, then the prediction made by the perceptron was wrong.  The
perceptron should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to reflect the
misprediction. In either case, weights are saturated at -16 or +15.

A secondary training mechanism also kicks in during demand fetches.  Before
the demand access triggers the next set of prefetches, Reject Table is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the underlying prefetch engine but rejected by the
perceptron filter.  Thus, the perceptron should have been more confident
about that particular prefetch.  Once such a scenario is identified, the state
of the execution at the time of prefetch is retrieved from the Reject Table.
The retrieved data is used to index into the various weights tables of the
perceptron and the corresponding values are updated by +1, saturating between
and -16 and +15.  This update reflects increased confidence for the prediction
corresponding to that prefetch.

This mechanism allows us to solve the classical problem of exploiting a lost
opportunity.  In prior perceptron based implementations (and in general,
prefetching algorithms), there is usually no way of knowing the result of not
prefetching a particular line.  Our two-step PPF architecture allows us to
overcome that issue.

@@

\item \textbf{Inferencing and Storing:} All prefetch recommendations to be
tested using the perceptron inferencing algorithm.  Perceptron output:
\textit{true} and \textit{false} should be saved appropriately, along with all
the metadata required for perceptron indexing.

\item \textbf{Retrieving and Training:} When the feedback is available, the
previously stored metadata can be used to re-index into the perceptron entries
and increment / decrement the weights.

\item \textbf{Feature Selection:} In Section~\ref{Impl-Features}, we show the features 
used for PPF. Six out of the nine features we developed
use information derived directly from program execution, irrespective of the baseline 
prefetcher. Beyond that, the feature set can be expanded to convey any
useful information from the prefetcher to the perceptron layer.  The 
methodology explained in Section~\ref{Method-Features} talks about developing a minimal 
feature set for PPF.
%For example, in our implementation ``Confidence,'' ``Current Signature XOR Delta,''
%and ``Page Address XOR Confidence'' are such features.

>>>>

In this section, we talk about the the general structure of the 
Perceptron-Based Prefetch Filtering (PPF) for enhancing an underlying prefetcher.  
PPF is a generalised prefetch filtering mechanism and can be adapted to 
most prefetchers with minimal modifications. %Gino: Are these modifications minimal?

\subsection{The Perceptron Filter}
\label{Arch-Perceptron}
The Perceptron Filter is organized as tables of perceptron weights, 
with a separate table of weights for each feature.  
A different number of bits are used for each feature to index into its 
corresponding table.
%Gino: repharse below?
Each feature's value is used to index into the corresponding table.
This indexing is done using a different number of bits for each feature.  
At most, 12 bits of a feature are used to index into the
weights table. Certain features require more resolution
\textit{i.e.}, the full 12 bits of indexing.  On the other hand, some
features requiring lower resolution require as few as 7 bits of indexing. 
The variable indexing was determined by studying the
features and fine-tuned empirically to achieve a good accuracy
vs hardware overhead trade-off.  Exact details are mentioned under ``Area Overheads''
in Section~\ref{Method-Overheads}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15 and initialized to zero at the 
beginning of a program's execution. Our proposed design uses nine features, 
represented in nine perceptron weight tables.

\textbf{Inference}\newline In the ChampSim simulator, the prefetching method is 
invoked on every L2 Cache demand access.  At that point, the base prefetcher
has the option to perform a prefetch -- if it does, then it has a choice
of how many cache lines to prefetch.  The suggested prefetches can be either placed
in the L2 cache or the last-level cache based on the prefetcher's internal confidence mechanism.

When the base prefetcher is triggered, it begins to suggest candidates for
prefetching. All suggested prefetches are passed through the perceptron
filter to decide if they are qualified for final prefetching. 
he perceptron decides whether to prefetch a candidate by looking at the microarchitectural state 
\em} i.e. the features, at that instant. Each feature is hashed to form an index into a table of up to 
4096 entries dedicated for that feature.

Once all the weights are retrieved, they are summed and compared to a preset threshold 
,PERC\_ THRESHOLD\_LO.  Only the prefetch candidates with perceptron sum higher than 
the threshold qualify for prefetching.  The prefetches that qualify through the perceptron 
stage are recorded in the ``Prefetch Table". The prefetch table is a 1024-entry, direct mapped structure taht
conatains all metadata required to recreate the features. 	%Gino: Recreate the features?
When the feedback of the current prefetched line is available at a later stage, the
stored data is used to train the perceptron.

The sum of the perceptron weights are also used to decide whether to place teh 
data in the L2 or the last level cache. All prefetch candidates that are qualified until
this stage are compared against PERC\_THRESHOLD\_HI to decide the fill
level. The two thresholds: PERC\_THRESHOLD\_LO and PERC \_THRESHOLD\_HI are
empirically set.

The perceptron sum can be considered the sum of the individual contribution
of a feature.  The value of each contribution corresponds to a feature's
confidence in the final decision on whether to prefetch a cache line. By summing the
individual contributions, the final perceptron sum denotes the overall
confidence for that prefetch suggestion.  By comparing the perceptron sum
to two different thresholds, we divide the confidence scale of the
prefetch suggestion into three bins. The first bin corresponds to the lowest
confidence and leads to prefetch candidate being rejected.  The next bin
corresponds to prefetches with a moderate confidence level.  This bin prefetches
 directeky to fill the bigger last level cache and prevents polluting
the smaller L2 Cache.  The highest confidence prefetches are filled in the
L2 cache.

In addition to the prefetch table mentioned above, PPF also maintains a
``Reject Table.''  The reject table is a 1024-entry deep direct-mapped
table.  If a prefetch suggestion is rejected by the perceptron layer, it is
logged into the reject table.  The filer is used to train the perceptron to %Gino: Filer?
avoid false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

@@

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 Cache.  This action triggers training of the base prefetcher as well as the
perceptron layer.  Training the perceptrons involves accessing
the metadata that was stored in Prefetch Table or the Reject Table. 
The cache lines address of the block being trained is used to access a table,
using 10-bits of the address for indexing and another 6-bits to perform tag matching.
Once the state of the program at the time of prefetching is available, it is 
%Gino: By "state of the program" are you referring to the state of the features?
used to index into the perceptron weights table.

If the demanded block that triggers the training was tagged as a valid
prefetch in the prefetch table, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1
if the predicted sum does not cross a pre-defined threshold. These training 
thresholds are introduced to avoid overfitting of the perceptron weights to the given program 
behaviour. These thresholds are referred to as POS\_UPDATE\_THRESHOLD ($\theta_p$) and 
NEG\_UPDATE\_THRESHOLD ($\theta_n$),  respectively for the positive and 
negative values of training saturation.

If a cache block eviction led to training and the corresponding valid entry is found in the
prefetch table without the block being accessed, then the prediction made by the perceptron was wrong.
The perceptron should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to reflect the
misprediction. In either case, weights are saturated at -16 or +15. %Gino: Describe how these limits were decided?

A secondary training mechanism also kicks in during demand fetches.  Before
the demand access triggers the next set of prefetches, the reject table is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the underlying prefetch engine, but rejected by the
perceptron filter.  Thus, the perceptron should have been more confident
about that particular prefetch.  Once such a scenario is identified, the state %Gino: state of execution == value of features?
of the execution at the time of prefetch is retrieved from the reject table.
The retrieved data is used to index into the various weights tables of the
perceptron and the corresponding values are updated by +1, saturating between
and -16 and +15.  This update reflects increased confidence for the prediction
corresponding to that prefetch.

This mechanism allows us to exploit a previously lost opportunity.  
In prior perceptron based implementations, and in general,
prefetching algorithms, there is usually no way of knowing the result of not
prefetching a particular line.  Our two-step PPF architecture allows us to
overcome that issue.

@@

\item \textbf{Inferencing and Storing:} All prefetch recommendations are
tested using the perceptron inferencing algorithm. The perceptron's output,
\textit{true} or \textit{false}, should be saved appropriately, along with all
 metadata required for perceptron indexing.

\item \textbf{Retrieving and Training:} When feedback for a prefetch is available, the
previously stored metadata can be used to re-index into the perceptron entries
and increment or decrement the weights.

\item \textbf{Feature Selection:} In Section~\ref{Impl-Features}, we show the features 
used for PPF. Six out of the nine features we developed
use information derived directly from program execution, irrespective of the baseline 
prefetcher. Beyond that, the feature set can be expanded to convey any
useful information from the prefetcher to the perceptron filter.  The 
methodology explained in Section~\ref{Method-Features} talks about developing a minimal 
feature set for PPF.
%For example, in our implementation ``Confidence,'' ``Current Signature XOR Delta,''
%and ``Page Address XOR Confidence'' are such features.

### 2018-11-29 design_impl.tex
Gino's first pass, modifying the abs, intro, design_arch, and design_impl

This section describes our implementation of PPF and the range of 
features that are ultimately used by the perceptron for correlating 
the prefetches.  For a practical implementation, we have used SPP 
as the base engine prefetcher.

@@

\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The original
  SPP would not consider the slots available in the L2 MSHR queue.  It had an
  assumption that at no time can the number of prefetches suggested can
  exceed the capacity of the queue.  While it was not proven in the paper, the
  internal confidence mechanism made sure that the above assumption was maintained.  
  In PPF since there can be no assumption on the degree of aggressiveness of the 
  underlying prefetcher, we explicitly check that at no point should the number 
  of suggested prefetches exceed the L2 MSHR queue.

\item \textbf{Enhanced Prefetch Table}\newline The Prefetch Table introduced in the 
  original SPP keeps a track of up to 1024 L2 cache prefetch suggestions.  
  For PPF, the Table was modified to store metadata such as program counter, 
  memory access address etc.  This metadata is needed to retrieve the state 
  of the features to train the perceptron at a later stage in time.

\item \textbf{Reject Table introduced}\newline The Prefetch Table mentioned
  above keeps track of the prefetches that actually happened in the L2 Cache.
  Correspondingly, we introduced the Reject Table to keep a track of the
  prefetches that were suggested by the base SPP engine but were rejected by the
  perceptron.  In case it is detected that a future demand fetch could have used
  this rejected prefetch, the perceptron weights are updated to reflect that
  misprediction.

@@

\begin{itemize}
\item \textbf{Base Address} of the demand access that triggered the
  prefetch.  Since prefetches are triggered from the L2 demand access,
  they tend to be correlated with the triggering address.

@@

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
  PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
  PC before the instruction that triggered the current prefetch.
  Hashing together the last three PCs tell PPF about the path
  that led to the current demand access and helps capture and
  branching information of the current basic block.  PCs are shifted
  in the increasing order of history before being hashed together.
  This is done to avoid the resultant value of zero in case 2 or more
  PCs are the same.  Additionally, obfuscating the information as it
  gets old allows us to get a wider and an approximate look into the
  program history.

@@

% djimenez: this is a little vague. can you be more precise?
\item \textbf{Current Signature XOR Delta}: Recall from
  the discussion of SPP in Section \ref{Background-SPP} that the new signature
  is generated using the old signature and the block delta.  While ``Current
  Signature XOR Delta'' is not the exact formula for generating the future
  signature, it gives an approximate idea of the path that the combination of
  these two values can lead to.

>>>>

This section describes our implementation of PPF and the range of 
features that are used by the perceptron for correlating 
the prefetches.  For a practical implementation, we have used SPP 
as the base engine prefetcher.

@@

\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The original
  SPP does not consider the slots available in the L2 MSHR queue.  It has the
  assumption that at no time can the number of suggested prefetches
  exceed the capacity of the queue.  While it was not described in the paper, the
  internal confidence mechanism makes sure that the above assumption was maintained.  
  In PPF, since there cannot be an assumption on the degree of aggressiveness of the 
  underlying prefetcher, we explicitly check that at no point should the number 
  of suggested prefetches exceed the L2 MSHR queue.

\item \textbf{Enhanced Prefetch Table}\newline The prefetch table introduced in the 
  original SPP keeps track of up to 1024 L2 cache prefetch suggestions.  
  For PPF, the prefetch table is modified to store metadata such as program counter, 
  memory access address etc.  This metadata is needed to retrieve the state  %Gino: What's the etc.?
  of the features to train the perceptron at a later stage in time. The prefetch table 
  mentioned above keeps track of the prefetches that actually happened in the L2 Cache.
  We introduce the reject table to keep a track of the
  prefetches that were suggested by the base SPP engine but were rejected by PPF.  
  In case it is detected that a future demand fetch could have used
  this rejected prefetch, the perceptron weights are updated to reflect that
  misprediction.

@@

\begin{itemize}
\item \textbf{Base Address} of the demand access that triggered the
  prefetch.  Since prefetches are triggered from the L2 demand access,
  they tend to correlate to the triggering address.

@@

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
  PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
  PC before the instruction that triggered the current prefetch.
  Hashing together the last three PCs tell PPF about the path
  that led to the current demand access and helps capture and
  branching information of the current basic block.  PCs are shifted
  in the increasing order of history before being hashed together.
  This is done to avoid the resultant value of zero when 2 or more
  PCs are the same.  Additionally, obfuscating the information as it
  gets old allows us to get a wider and approximate look into the
  program's history.

@@

% djimenez: this is a little vague. can you be more precise?
\item \textbf{Current Signature XOR Delta}: Recall from
  the discussion of SPP in Section \ref{Background-SPP} that the new signature
  is generated using the old signature and the current block delta.  
%Gino: Added the sentence below, is this correct?
  The result of this feature is the next delta that is predicted to be accessed based on the 
  delta predicted by SPP. While ``Current Signature XOR Delta'' is not the exact formula for generating the future
  signature, it gives an approximate idea of the path that the combination of
  these two values can lead to.

### 2018-11-29 intro.tex
Gino's first pass, modifying the abs, intro, design_arch, and design_impl

\section{Introduction}
\label{Introduction}
Processor scaling and memory scaling have been enabled with different
philosophies in mind.  While processor scaling focused on speed improvements,
memory scaling was centred around increasing the storage capacity.  Process
technology reaching scaling limitations led to the Memory Wall~\cite{MemWall}
-- the increasing performance gap between processor speed and the memory
speeds.  Various techniques have been developed to combat this wide gap. Data
prefetching is one such important technique. Data prefetching exploits the
fact that in most applications, memory accesses have a well-defined pattern.

An ideal prefetching scheme would capture these memory access patterns to
predict future accesses in a timely manner.  Other important decisions faced
by the prefetcher are when to bring in the prefetch block of data, which level
of the cache hierarchy to place the block in, and what block to evict to
accommodate the incoming block.  Memory accesses patterns can be as simple as
fetching the next cache line or a complex pattern such as pointer chasing.
Predicting future accesses requires a trade-off between coverage and accuracy.
Most prefetchers maintain an internal confidence mechanism by keeping counters
to track events such as cache hits / misses, references to an entry in the
table, etc. By thresholding this confidence against different values, the
degree of aggressiveness can be adjusted.

For a prefetcher to capture highly irregular access patterns, it needs to be
highly aggressive (high coverage).  Thus, the prefetcher would recommend many
useless prefetches, leading to cache pollution (low accuracy).  This effect is
even more pronounced in a multi-core scenario where the last-level cache is a
shared resource~\cite{Friendly}.  Conversely, a conservative prefetcher would
be highly accurate but might not make enough useful prefetches.

We propose Perceptron-based Prefetch Filtering (PPF).  It is an enhancement to
the existing state of the art prefetcher and robustly overcomes the coverage
vs.  accuracy trade-off.  The idea is to use a state-of-the-art prefetcher as
the base engine.  For that we choose Signature Path Prefetcher (SPP) and
modify the design to make it as aggressive as possible.  This modification
enables the base engine to capture complex memory access patterns and go
deeper into the speculation path, leading to increased coverage. Normally such
aggressive prefetching would come at the cost of increased DRAM traffic.
However, prefetch suggestions recommended by the base prefetch engine are
passed through the perceptron based filter.  Over time, perceptron layer
learns to correlate the prefetch recommendations with various available
features, leading to elimination of useless prefetches. This filtering leads
to overall increased accuracy of the prefetch scheme and reduction in DRAM
traffic.

@@

\item An on-line neural model is used for hardware data
  prefetching.  Previous work in this area either relied on program
  semantics~\cite{Semantics} or were application specific~\cite{Datacenter}.
% djimenez: removing reference to this arXiv paper. we are not obligated to cite it.
%, or
%  could not yield a significant improvement over the baseline prefetcher
%  used~\cite{BadPerc}.

\item Implementing the perceptron filter in a robust and accurate
  practical prefetching mechanism, yielding a significant performance
  improvement as compared to the other state of the art prefetchers.
  Moreover, the perceptron learns to adapt itself to shared resource
  constraints, leading to further increased performance in multi-core and
  lower capacity environments.

\item Defining a methodology for determining an appropriate set of 
  features for prediction regardless of the underlying prefetcher used.
  More details are explained in Section \ref{Method-Features}.

@@

On a single core configuration, PPF increased performance by 6.84\% as
compared to the next best prefetcher, SPP.  On a multi-core system running a
mix of memory intensive subset of the SPEC 2017 traces, PPF saw an improvement
of 11.9\% over SPP for a 4-core system and 10.67\% for an 8-core system.

>>>>

\section{Introduction}
\label{Introduction}
Processor scaling and memory technologies have been developed with different
philosophies in mind.  While processor scaling focused on speed improvements,
memory scaling was centred around increasing the storage capacity.  The
difference in each technology's scaling has led to the Memory Wall~\cite{MemWall}
-- the increasing performance gap between processor and memory
speeds. Data prefetching is one important technique that has been developed 
to bridge this widening performance gap. Data prefetching exploits the fact that in most applications,
 memory accesses have a repeating and predictable pattern to speculatively fetch useful data from slower levels
of the cache hierarchy into faster higher levels of cache. %phrasing?

An ideal prefetching scheme would perfectly capture memory access patterns to
predict and request future accesses in a timely manner. Other important decisions faced
by the prefetcher are when to bring in the prefetch block of data, which level
of the cache hierarchy to place the block in, and what block to evict to
accommodate the incoming block. Memory accesses patterns can be as simple as
fetching the next cache line or a complex pattern such as pointer chasing.
Predicting future accesses requires a trade-off between coverage, the percent of predictions that 
prevented a cache miss, and accuracy, the percent of predictions made that were correct. 
%Gino: Just added a short explanation of coverage and accuracy tradeoff, not sure if necessary
A prefetcher may have high coverage by reducing the number of misses by simply requesting a large 
amount of data, however this does not imply accuracy. Even if the prefetcher brought in data accessed in the future,
a high percentage of the data requested could never be accessed meaning the prefetcher had low accuracy, 
resulting in wasted resources as well as cache pollution. Likewise, a prefetcher can have a 
higher accuracy by being completely sure the data requested is used in the future, 
but may not affect performance since it did not have the coverage needed to make an impact. %end addition.
Most prefetchers maintain an internal confidence mechanism by keeping counters
to track events such as cache hits, misses, references to an entry in the prefetcher's
structure, etc. By comparing the confidence indicated by the counters against different 
threshold values, the aggressiveness of the prefetcher can be adjusted, resulting in lower
coverage and higher accuracy.

For a prefetcher to capture highly irregular access patterns, it needs to be
highly aggressive with high coverage. This causes the prefetcher to recommend many
low accurracy prefetches that lead to the cache being polluted with useless data.  
This effect is more pronounced in a multi-core scenario where the last-level cache is a
shared resource~\cite{Friendly}.  Conversely, a conservative prefetcher would
be highly accurate but might not make enough useful prefetches due to its lower coverage.

We propose Perceptron-based Prefetch Filtering (PPF) as an enhancement to
the existing state of the art prefetchers to robustly overcome the coverage
vs. accuracy trade-off.  The idea is to use a state-of-the-art prefetcher as
the base engine. For that we choose Signature Path Prefetcher (SPP) and
modify the design to make it as aggressive as possible.  This modification
enables the base engine to capture complex memory access patterns and go
deeper into the speculation path, increasing coverage without sacrificing accuracy. 
Normally such aggressive prefetching would come at the cost of increased DRAM traffic
and cache pollution. However, prefetch suggestions recommended by the base prefetch
engine are passed through the perceptron based filter to ensure an accurate prediction.
Over time, PPF learns to correlate the prefetch recommendations with various available
features, leading to elimination of useless prefetches. This filtering leads
to overall increased accuracy of the prefetch scheme and reduction in DRAM
traffic and cache pollution.

@@

\item An on-line neural model used for hardware data
  prefetching.  Previous work in this area either relied on program
  semantics~\cite{Semantics} or were application specific~\cite{Datacenter}.
% djimenez: removing reference to this arXiv paper. we are not obligated to cite it.
%, or
%  could not yield a significant improvement over the baseline prefetcher
%  used~\cite{BadPerc}.

\item Implementing the perceptron filter in a robust and accurate
  practical prefetching mechanism, yielding a significant performance
  improvement compared to other state of the art prefetchers.
  Moreover, the perceptron learns to adapt itself to shared resource
  constraints, leading to further increased performance in multi-core and
  lower memory capacity environments.

\item A methodology for determining an appropriate set of 
  features for prediction regardless of the underlying prefetcher used.
  More details are explained in Section \ref{Method-Features}.

@@

On a single core configuration, PPF increases performance by 6.84\%
compared to the baseline prefetcher, SPP.  On a multi-core system running a
mix of memory intensive subset of the SPEC 2017 traces, PPF saw an improvement
of 11.9\% over SPP for a 4-core system and 10.67\% for an 8-core system.
